---
title: "Decision Tree Challenge"
subtitle: "Feature Importance and Categorical Variable Encoding"
format:
  html: 
    toc: true
    toc-depth: 3
    toc-title: "Contents"
execute:
  echo: false
  eval: true
---

# ðŸŒ³ Decision Tree Challenge - Feature Importance and Variable Encoding [#](#sec-discussion)

Let's assume we want to predict house prices and understand which features matter most. The key question is: **How does encoding categorical variables as numbers affect our understanding of feature importance?**

## The Ames Housing Dataset ðŸ 

We are analyzing the Ames Housing dataset which contains detailed information about residential properties sold in Ames, Iowa from 2006 to 2010. This dataset is perfect for our analysis because it contains a categorical variable (like zip code) and numerical variables (like square footage, year built, number of bedrooms).

## The Problem: ZipCode as Numerical vs Categorical

**Key Question:** What happens when we treat zipCode as a numerical variable in a decision tree? How does this affect feature importance interpretation?

**The Issue:** Zip codes (50010, 50011, 50012, 50013) are categorical variables representing discrete geographic areas, i.e. neighborhoods. When treated as numerical, the tree might split on "zipCode > 50012.5" - which has no meaningful interpretation for house prices.  Zip codes are non-ordinal categorical variables meaning they have no inherent order that aids house price prediction (i.e. zip code 99999 is not the priceiest zip code).

## Data Loading and Model Building

```{r}
#| label: load-and-model-r
#| echo: false
#| message: false
#| warning: false

# Load libraries
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(rpart))
if (!require(rpart.plot, quietly = TRUE)) {
  install.packages("rpart.plot", repos = "https://cran.rstudio.com/")
  library(rpart.plot)
}

# Load data
sales_data <- read.csv("https://raw.githubusercontent.com/flyaflya/buad442Fall2025/refs/heads/main/datasets/salesPriceData.csv")

# Prepare model data (treating zipCode as numerical)
model_data <- sales_data %>%
  select(SalePrice, LotArea, YearBuilt, GrLivArea, FullBath, HalfBath, 
         BedroomAbvGr, TotRmsAbvGrd, GarageCars, zipCode) %>%
  na.omit()

# Split data
set.seed(123)
train_indices <- sample(1:nrow(model_data), 0.8 * nrow(model_data))
train_data <- model_data[train_indices, ]
test_data <- model_data[-train_indices, ]

# Build decision tree
tree_model <- rpart(SalePrice ~ ., 
                    data = train_data,
                    method = "anova",
                    control = rpart.control(maxdepth = 3, 
                                          minsplit = 20, 
                                          minbucket = 10))

cat("Model built with", sum(tree_model$frame$var == "<leaf>"), "terminal nodes\n")
```

## Tree Visualization

```{r}
#| label: visualize-tree-r
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

# Visualize tree
if (require(rpart.plot, quietly = TRUE)) {
  rpart.plot(tree_model, 
             type = 2,
             extra = 101,
             fallen.leaves = TRUE,
             digits = 0,
             cex = 0.8,
             main = "Decision Tree (zipCode as Numerical)")
} else {
  plot(tree_model, uniform = TRUE, main = "Decision Tree (zipCode as Numerical)")
  text(tree_model, use.n = TRUE, all = TRUE, cex = 0.8)
}
```

## Feature Importance Analysis

```{r}
#| label: feature-importance-r
#| echo: false
#| message: false
#| warning: false

# Extract and display feature importance
importance_df <- data.frame(
  Feature = names(tree_model$variable.importance),
  Importance = as.numeric(tree_model$variable.importance)
) %>%
  arrange(desc(Importance)) %>%
  mutate(Importance_Percent = round(Importance / sum(Importance) * 100, 2))

# Check zipCode ranking
zipcode_rank <- which(importance_df$Feature == "zipCode")
zipcode_importance <- importance_df$Importance_Percent[zipcode_rank]
```

```{r}
#| label: importance-plot-r
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 5

# Plot feature importance
library(ggplot2)
ggplot(importance_df, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  coord_flip() +
  labs(title = "Feature Importance (zipCode as Numerical)",
       x = "Features", y = "Importance Score") +
  theme_minimal()
```

## Critical Analysis: The Encoding Problem

::: {.callout-warning}
## âš ï¸ The Problem Revealed

**What to note:** Our decision tree treated `zipCode` as a numerical variable. This leads to zip code being unimportant. Not surprisingly, because there is no reason to believe allowing splits like "zipCode < 50012.5" should be beneficial for house price prediction. This false coding of a variable creates several problems:

1. **Potentially Meaningless Splits:** A zip code of 50013 is not "greater than" 50012 in any meaningful way for house prices
2. **False Importance:** The algorithm assigns importance to zipCode based on numerical splits rather than categorical distinctions OR the importance of zip code is completely missed as numerical ordering has no inherent relationship to house prices.
3. **Misleading Interpretations:** We might conclude zipCode is not important when our intuition tells us it should be important (listen to your intuition).

**The Real Issue:** Zip codes are categorical variables representing discrete geographic areas. The numerical values have no inherent order or magnitude relationship to house prices. These must be modelled as categorical variables.
:::

## Proper Categorical Encoding: The Solution

Now let's repeat the analysis with zipCode properly encoded as categorical variables to see the difference.

### Categorical Encoding Analysis

```{r}
#| label: categorical-r
#| echo: false
#| message: false
#| warning: false

# Convert zipCode to factor (categorical)
model_data_cat <- model_data %>%
  mutate(zipCode = as.factor(zipCode))

# Split data
set.seed(123)
train_indices_cat <- sample(1:nrow(model_data_cat), 0.8 * nrow(model_data_cat))
train_data_cat <- model_data_cat[train_indices_cat, ]
test_data_cat <- model_data_cat[-train_indices_cat, ]

# Build decision tree with categorical zipCode
tree_model_cat <- rpart(SalePrice ~ ., 
                        data = train_data_cat,
                        method = "anova",
                        control = rpart.control(maxdepth = 3, 
                                              minsplit = 20, 
                                              minbucket = 10))

# Feature importance with categorical zipCode
importance_cat <- data.frame(
  Feature = names(tree_model_cat$variable.importance),
  Importance = as.numeric(tree_model_cat$variable.importance)
) %>%
  arrange(desc(Importance)) %>%
  mutate(Importance_Percent = round(Importance / sum(Importance) * 100, 2))

# Check if zipCode appears in tree
zipcode_in_tree <- "zipCode" %in% names(tree_model_cat$variable.importance)
if(zipcode_in_tree) {
  zipcode_rank_cat <- which(importance_cat$Feature == "zipCode")
}
```

### Tree Visualization: Categorical zipCode

```{r}
#| label: visualize-tree-cat-r
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

# Visualize tree with categorical zipCode
if (require(rpart.plot, quietly = TRUE)) {
  rpart.plot(tree_model_cat, 
             type = 2,
             extra = 101,
             fallen.leaves = TRUE,
             digits = 0,
             cex = 0.8,
             main = "Decision Tree (zipCode as Categorical)")
} else {
  plot(tree_model_cat, uniform = TRUE, main = "Decision Tree (zipCode as Categorical)")
  text(tree_model_cat, use.n = TRUE, all = TRUE, cex = 0.8)
}
```

### Feature Importance: Categorical zipCode

```{r}
#| label: importance-plot-cat-r
#| echo: false
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 5

# Plot feature importance for categorical zipCode
library(ggplot2)
ggplot(importance_cat, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col(fill = "darkgreen", alpha = 0.7) +
  coord_flip() +
  labs(title = "Feature Importance (zipCode as Categorical)",
       x = "Features", y = "Importance Score") +
  theme_minimal()
```


## Discussion Questions for Challenge {#sec-discussion}


1. **Numerical vs Categorical Encoding:** There are four models above, two in R and two in Python. For each language, the models differ by how zip code is modelled, either as a numerical variable or as a categorical variable. Given what you know about zip codes and real estate prices, how should zip code be modelled, numerically or categorically?

### Given what I know about zip codes and real estate prices, zip code should be modelled as categorical because when the zipcode is treated as numerical, the tree ends up splitting for example "zipcode >50012.5". This does not make sense because zip code of 50013 is not "greater than" 50012 in terms of house prices. Logially, zip codes do play a significant role in prediciting house prices as zip codes with more income tend to have houses that are worth more. On the other hand, the right way to model zip code would be categorical where we treat it as discrete cateogies. When zipcode is numerical, the feature importance ranks zipcode very low whereas when zipcode is categorical, there is high importance as it becomes one of the most important features. Moreover, when zip codes are treated as categories, the decision tree can split on "zipcode=-50010" vs "zipcode==50010" and make clear distinctions on different neightborhoods. In regards to real estate prices, it would be a wrong business decision to use numerical encoding since you would conclude location does not matter for house prices. Therefore, data scientists should be very careful on how they prepare their data as it can lead to a common mistake that makes wrong business insights.

2. **R vs Python Implementation Differences:** When modelling zip code as a categorical variable, the output tree and feature importance differs quite significantly between R and Python. Investigate why this is the case. Which language would you say does a better job of modelling zip code as a categorical variable? Why is this the case? Do you see any documentation suggesting the other language does a better job? If so, please provide a quote from the documentation.

### R and Python differ in categorical variable handling because R's rpart package supports categorical variables and has an algorithm that splits directly on categorical vairable and treates it as a single feature for importance calculation. On the other hand, Python's scikit-learn requires all input data to be numerical and must use one-hot encoding, which creates seperaate binary columns for each zip code. Therefore the importance is spread across multiple dummy variables. R does a significantly better job of modeling categorical variables because R creates meaningful splits that group related cateogries together but Python creates artificial splits on individual dummy variables that are harder to interpret. We can see in feature importance that Python dilutes importance across multiple dummy variables which is underestimating the true importance. R maintains reasonable tree complexity with logical categorical splits whereas Python creates overly complex trees with many splits on dummy variables. To support my claim, I will provide a quote from the documentation of the sklearn.tree.DecisionTreeRegression site. The quote says, "However, the scikit-learn implementation does not support categorical variables for now." This explains why Python does a worse job of modelling categorical variables.




**What makes a great report:**

- **Clear narrative:** Tell the story of what you discovered about decision tree feature importance
- **Insightful analysis:** Focus on the most interesting differences between numerical and categorical encoding
- **Professional presentation:** Clean, readable, and engaging
- **Concise conclusions:** No AI babble or unnecessary technical jargon
- **Human insights:** Your interpretation of what the feature importance rankings actually mean (or don't mean)
- **Documentation-based analysis:** For question 2, ground your analysis in actual library documentation

**What we're looking for:** A compelling 1-2 minute read that demonstrates both the power of decision trees for interpretability and the critical importance of proper variable encoding.
:::

### Questions to Answer for 75% Grade on Challenge

1. **Numerical vs Categorical Analysis:** Provide a clear, well-reasoned answer to question 1 about how zip codes should be modelled. Your answer should demonstrate understanding of why categorical variables need special treatment in decision trees.

### Questions to Answer for 85% Grade on Challenge

2. **R vs Python Implementation Analysis:** Provide a thorough analysis of question 2, including investigation of the official documentation for both `rpart` (R) and `sklearn.tree.DecisionTreeRegressor` (Python). Your analysis should explain the technical differences and provide a reasoned opinion about which implementation handles categorical variables better.

### Questions to Answer for 95% Grade on Challenge

3. **Professional Presentation:** Your discussion answers should be written in a professional, engaging style that would be appropriate for a business audience. Avoid technical jargon and focus on practical implications.  disc

### Questions to Answer for 100% Grade on Challenge

4. **Documentation Integration:** For question 2, include a specific quote from the official documentation of `sklearn.tree.DecisionTreeRegressor` that supports your analysis. 
